\documentclass{mynotes}

\usepackage{etex} \usepackage{mymathdn} \usepackage{mycodesupport}
\usepackage{mygeneralsettings} \usepackage{mymathntn}



% \input{gitInfo}
\lstset{language=Matlab}

\newcommand{\by}{\mathbf{Y}} %
\newcommand{\bx}{\mathbf{X}} %
\newcommand{\bt}{\mathbf{T}} %
\newcommand{\bg}{\mathbf{G}} %
\newcommand{\bz}{\mathbf{Z}} %
\newcommand{\bp}{\mathbf{P}} %
\newcommand{\br}{\mathbf{R}} %
\newcommand{\bw}{\mathbf{W}} %
\newcommand{\eye}{\mathbf{I}} %

\newcommand{\Hm}[1]{\mathbf{H}_{#1}} %hat matrix
\newcommand{\Dm}[1]{\mathbf{D}_{#1}} %diagonal matrix
\newcommand{\Mm}[1]{\mathbf{M}_{#1}} %annihilator matrix

% random effects
\newcommand{\Qs}{Q_{\mathcal{S}}} % Q_S
\newcommand{\hQs}{\hat{Q}_{\mathcal{S}}} % \hat{Q}_S
\newcommand{\Qt}{Q_{\mathcal{T}}} % Q_T
\newcommand{\h}[2]{\hat{#1}_{\text{\an{#2}}}} %

\title{\texttt{ivreg}}%
\author{Michal
  Kolesár\thanks{\href{mailto:kolesarmi@googlemail.com}{kolesarmi@googlemail.com}}}%
% \date{Version \gitTag, \gitDate\thanks{Version hash \gitHash, compiled \today
%   }}%

\begin{document}
\maketitle

\section{Model and notation}
The basic model is:
\begin{equation}\label{eq:1}
  y_{i}=T_{i}\beta+W_{i}'\psi+\epsilon_{i},
\end{equation}
where $y_{i}\in\mathbb{R}$ is the outcome variable, $T_{i}\in\mathbb{R}$ is a
single endogenous regressor, $W_{i}\in\mathbb{R}^{L}$ is a vector of exogenous
regressors (covariates), and $\epsilon_{i}$ is a structural error. The
identifying assumption is that the structural error $\epsilon_{i}$ is
uncorrelated with the covariates $W_{i}$ and a given vector of instruments
$Z_{i}\in\mathbb{R}^{K}$:
\begin{equation}
  \label{eq:id}
  \E\left[\epsilon_{i}
    \begin{pmatrix}
      Z_{i}\\W_{i}
    \end{pmatrix}\right]
  =0.
\end{equation}
We observe an i.i.d.~sample $\{Y_{i},T_{i},W_{i},Z_{i}\}_{i=1}^{n}$. The
arguments of \lstinline!ivreg! are the matrices $\by$, $\bt$, $\bz$, and $\bw$,
with rows $y_{i}$, $T_{i}$, $W_{i}'$ and $Z_{i}'$.

For any full-rank $n\times m$ matrix $\mathbf{A}$, let
$\Hm{\mathbf{A}}=\mathbf{A}(\mathbf{A}'\mathbf{A})^{-1}\mathbf{A}'$ denote the
associated $n\times n$ projection matrix (also known as the hat matrix), and let
$\Dm{\mathbf{A}}$ be an $n\times n$ diagonal matrix with
$(\Hm{\mathbf{A}})_{ii}$ on the diagonal. Let $\eye_{m}$ denote the $m\times m$
identity matrix, and let $\Mm{\mathbf{A}}=\eye_{n}-\Hm{\mathbf{A}}$ denote the
annihilator matrix. Let $\mathbf{A}_{\perp}=\Mm{\bw}\mathbf{A}$ denote the
residual from the sample projection of $\mathbf{A}$ onto $\bw$.

\section{Estimation}
The command
\begin{lstlisting}
betahat = ivreg(Y, T, Z, W);
\end{lstlisting}
returns a vector \lstinline!betahat! of different estimators of the causal
effect $\beta$ in Equation (\ref{eq:1}). In particular:
\begin{lstlisting}
betahat = [ols, tsls, liml, mbtsls, jive, ujive, rtsls]
\end{lstlisting}
where \texttt{ols} is the least-squares estimator, \texttt{tsls} is the
two-stage least squares estimator, \texttt{liml} is the limited information
maximum likelihood estimator \citep{ar49}, \texttt{mbtsls} is the modified
bias-corrected two-stage least squares estimator \citep{anatolyev11,kcfgi11},
\texttt{jive} is the jackknife iv estimator \citep{ph77,aik99}, also called
JIVE1, \texttt{ujive} is the \citet{kolesar12late} version of the jackknife iv
estimator, and \texttt{rtsls} is the reverse two-stage least squares estimator.

\paragraph{Expressions for estimators of $\beta$}
All estimators can be written as
\begin{equation}\label{eq:hatbeta}
  \hat{\beta}=\frac{\hat{\bp}'\by}{\hat{\bp}'\bt},
\end{equation}
where the form of $\hat{\bp}$ depends on the estimator. In particular,
\begin{align*}
  \hat{\bp}_{\mathtt{ols}}&=\bt_{\perp}=(\eye-0\cdot \Mm{\bz_{\perp}})\bt_{\perp} ,\\
  \hat{\bp}_{\mathtt{tsls}}&=\Hm{\bz_{\perp}}\bt=(\eye-1\cdot \Mm{\bz_{\perp}})\bt_{\perp},\\
  \hat{\bp}_{\mathtt{liml}}&=\Mm{\bw}(\eye-k_{\mathtt{liml}}\Mm{\bw,\bz})\bt=
  (\eye-k_{\mathtt{liml}}\Mm{\bz_{\perp}})\bt_{\perp} ,\\
  \hat{\bp}_{\mathtt{mbtsls}}&=\Mm{\bw}(\eye-k_{\mathtt{mbtsls}}\Mm{\bw,\bz})\bt=
  (\eye-k_{\mathtt{mbtsls}}\Mm{\bz_{\perp}})\bt_{\perp},\\
  \hat{\bp}_{\mathtt{jive}}&=\Mm{\bw}(\eye_{n}-\Dm{\bz,\bw})^{-1}(\Hm{\bz,\bw}-\Dm{\bz,\bw})\bt=
  \Mm{\bw}(\eye_{n}-(\eye_{n}-\Dm{(\bz,\bw)})^{-1}\Mm{(\bz,\bw)})\bt,\\
  \hat{\bp}_{\mathtt{ujive}}&=\left[(\eye_{n}-\Dm{(\bz,\bw)})^{-1}(\Hm{(\bz,\bw)}-\Dm{(\bz,\bw)})
    -(\eye_{n}-\Dm{\bw})^{-1}(\Hm{\bw}-\Dm{\bw})\right]\bt,\\
  \hat{\bp}_{\mathtt{rtsls}}&=\Hm{\bz_{\perp}}\by,
\end{align*}
where
\begin{align}\label{eq:k-liml}
  k_{\mathtt{liml}}&=\min\eig \left[\left( \begin{pmatrix} \by&\bt
      \end{pmatrix}'\Mm{\bz,\bw} \begin{pmatrix} \by&\bt
      \end{pmatrix}\right)^{-1}
    \begin{pmatrix}
      \by&\bt
    \end{pmatrix}' \Mm{\bw}\begin{pmatrix} \by&\bt
    \end{pmatrix}\right],
\end{align}
and $k_{\mathtt{mbtsls}}=(1-L/n)/(1-(K - 1)/n-L/n)$. The version of
\texttt{mbtsls} proposed in \citet{kcfgi11} uses
$k_{\mathtt{mbtsls}}=(1-L/n)/(1-K/n-L/n)$---the modification used here ensures
that when $K=1$ (there is a single instrument), the \texttt{tsls} and
\texttt{mbtsls} estimators will coincide.



\section{Standard Errors}
The command
\begin{lstlisting}
[betahat, se] = ivreg(Y, T, Z, W);
\end{lstlisting}
returns an estimate of standard errors for the different estimators of $\beta$.
\lstinline!se! is a $4\times 7$ matrix with rows computed as follows
\begin{itemize}
\item The first row gives estimates of the asymptotic standard errors under
  homoscedasticity ($\E[\epsilon_{i}^{2}\mid z_{i},w_{i}]=\E[\epsilon_{i}^{2}]$)
  and standard asymptotics, which hold the distribution of the observed data
  $(Y_{i},T_{i},W_{i},Z_{i})$ fixed as $n\to \infty$;
\item The second row gives heteroscedasticity-robust standard errors;
\item The third row gives standard errors which are valid under many-instrument
  asymptotics. Like in \citet{bekker94}, these errors are valid when the number
  of instruments, $K$, is allowed to grow in proportion with the sample size,
  $K/n\to \kappa$. In addition, the number of exogenous covariates, $W_{i}$, is
  also allowed to grow with the sample size, $L/n\to \lambda$, as in
  \citet{anatolyev11} and \citet{cfhssy11}. As in \citet{bekker94}, the
  structural error $\epsilon_{i}$ is assumed to be Normally distributed. Since
  \texttt{tsls} and \texttt{jive} are not consistent under these asymptotics,
  elements of the \lstinline!se! matrix that correspond to them return
  \lstinline!NaN!;
\item The fourth row gives standard errors that are valid under the many invalid
  instrument asymptotic sequence of \citet{kcfgi11}. Only \texttt{mbtsls} and
  \texttt{ujive} are consistent for $\beta$ under this sequence, so element of
  the \lstinline!se! matrix that correspond to other estimators return
  \lstinline!NaN!.
\end{itemize}


\paragraph{Standard errors under homoscedasticity and standard asymptotics}
Let
\begin{equation}\label{eq:sigeps}
  \hat{\sigma}^{2}_{\epsilon}(\hat{\beta})
  =  (\by_{\perp}-\bt_{\perp}\hat{\beta})'(\by_{\perp}
  -\bt_{\perp}\hat{\beta})/n.
\end{equation}
The standard error for \texttt{ols} is given by
\begin{equation*}
  \widehat{se}(\hat{\beta}_{\mathtt{ols}})=\sqrt{\frac{  \hat{\sigma}^{2}_{\epsilon}(\hat{\beta})}{
      \bt_{\perp}'\bt_{\perp}}}.
\end{equation*}
The estimator of the standard error of $\hat{\beta}$ for \texttt{tsls},
\texttt{mbtsls} and \texttt{liml} is given by
\begin{equation*}
  \widehat{se}(\hat{\beta})=\sqrt{\frac{  \hat{\sigma}^{2}_{\epsilon}(\hat{\beta})}{
      \bt'\Hm{\bz_{\perp}}\bt}}.
\end{equation*}
The standard errors will be different if they produce different estimates of
$\beta$. These expressions correspond to the standard errors for \texttt{tsls}
in Stata, as well as to the standard errors for \texttt{mbtsls} and
\texttt{liml} when the option ``\lstinline!coviv!'' is given \citep{bss07}. The
standard error for \texttt{jive} and \texttt{ujive} is given by:
\begin{equation*}
  \widehat{se}(\hat{\beta})
  =\frac{\sqrt{
      \hat{\sigma}^{2}_{\epsilon}(\hat{\beta})
      \cdot \hat{\bp}'\hat{\bp}}}{
      \hat{\bp}'\bt}.
\end{equation*}
This estimator matches the standard error estimator for \texttt{jive} in
Stata.\footnote{The \citet{aik99} JIVE1 esimator is referred to as UJIVE1 in
  Stata, and is computed by the \lstinline!jive! command with the
  \lstinline!ujive1! option, which is also the default option.}

% To motivate the standard error estimator for \texttt{rtsls}, note first that
% by arguments similar to deriving the \texttt{tsls} standard error estimator, a
% consistent estimator for the variance of $1/\hat{\beta}_{\mathtt{rtsls}}$ is
% given by
% \begin{equation*}
%   \hat{V}_{1/\hat{\beta}_{\mathtt{rtsls}}}=
%   \frac{  \hat{\nu}_{\hat{\beta}_{\mathtt{rtsls}}}'  \hat{\nu}_{\hat{\beta}_{\mathtt{rtsls}}}/n}{
%     \by'\Hm{\bz_{\perp}}\by},
% \end{equation*}
% where
% $\hat{\nu}_{\hat{\beta}_{\mathtt{rtsls}}}=\bt_{\perp}-\by_{\perp}(1/\hat{\beta}_{\mathtt{rtsls}})$.
% Therefore, by the delta method, a consistent estimator of the standard error
% of \texttt{rtsls} is given by
% \begin{equation*}
%   \widehat{se}(\hat{\beta}_{\mathtt{jive}})=\sqrt{  \frac{  \hat{\nu}_{\hat{\beta}_{\mathtt{rtsls}}}'  \hat{\nu}_{\hat{\beta}_{\mathtt{rtsls}}}/n}{
%       \by'\Hm{\bz_{\perp}}\by}\hat{\beta}_{\mathtt{rtsls}}^{4}.
%   }
% \end{equation*}


\paragraph{Standard errors under heteroscedasticity}
Let
\begin{equation*}
  \hat{\epsilon}_{\hat{\beta}}=\by-\bt\hat{\beta}-\bw\hat{\psi}(\hat{\beta})
  =\by_{\perp}-\bt_{\perp}\hat{\beta}
\end{equation*}
be an estimate of the structural error in Equation \eqref{eq:1}, where
$\hat{\psi}(\hat{\beta})=(\bw'\bw)^{-1}\bw(\by-\bt\hat{\beta})$. The
heteroscedasticity-robust standard error is given by
\begin{equation*}
  \widehat{se}(\hat{\beta})  =\frac{\sqrt{\sum_{i=1}^{n}
      \hat{\epsilon}_{\hat{\beta},i}^{2}\hat{\bp}_{i}^{2}}}{\hat{\bp}'\bt},
\end{equation*}
where for \texttt{ols}, $\hat{\bp}=\bt_{\perp}$, for \texttt{tsls},
\texttt{liml}, and \texttt{mjive}, $\hat{\bp}=\Hm{\bz_{\perp}}\bt$, for
\texttt{jive} $\hat{\bp}=\hat{\bp}_{\mathtt{jive}}$, and for \texttt{ujive},
$\hat{\bp}=\hat{\bp}_{\mathtt{ujive}}$.

\paragraph{Standard errors under many instrument asymptotics}
The estimator of the standard error for \texttt{liml} corresponds to
$\sqrt{-\h{\mathcal{H}}{re}^{11}}$, where $\h{\mathcal{H}}{re}^{11}$ is the
(1,1) element of the inverse Hessian of the random effects likelihood
\citep{kolesar12re},
    \begin{equation*}
      \h{\mathcal{H}}{re}^{11}= \frac{\hat{b}'\h{\Omega}{re}\hat{b}
        (\h{\lambda}{re}+K/n)}{n\h{\lambda}{re}}\left(
        \hQs\h{\Omega}{re,22}-S_{22}+\frac{\hat c}{1-\hat c}\frac{\hQs}{\hat{a}'
\h{\Omega}{re}^{-1}\hat{a}} \right)^{-1},
    \end{equation*}
where
    \begin{align*}
      S_{\perp}&=      \begin{pmatrix}
        \by&\bt
      \end{pmatrix}'
\Mm{\bz,\bw}
      \begin{pmatrix}
        \by&\bt
      \end{pmatrix}
/(n-K-L),&S&= \begin{pmatrix}
        \by&\bt
      \end{pmatrix}'
\Hm{\bz_{\perp}}
      \begin{pmatrix}
        \by&\bt
      \end{pmatrix}/n,\\
      \h{\lambda}{re}&=\max\eig(S^{-1}_{\perp}S)-K /n,\\
      \h{\Omega}{re}&=\frac{n-K-L}{n-L}S_{\perp}+\frac{n}{n-L}\left(S-
        \frac{\h{\lambda}{re}}{\hat{a}'S_{\perp}^{-1}\hat{a} } \hat{a}\hat{a}'\right),&
      \hat{a}&=
      \begin{pmatrix}
        \hat{\beta}_{\mathtt{liml}}\\1
      \end{pmatrix},
      \\
      \hQs &=\frac{\hat{b}'S \hat{b}}{\hat{b}'\h{\Omega}{re} \hat{b}},& \hat{b}
      &=
  \begin{pmatrix}
    1\\-\hat{\beta}_{\mathtt{liml}}
  \end{pmatrix},\\
\hat c&=\frac{\h{\lambda}{re} \hQs}{(K/n+\h{\lambda}{re})(1-L/n)}.
    \end{align*}
    The estimator for the standard error of \texttt{mbtsls} is based on the
    Hessian of the uncorrelated random effects likelihood with the estimator of
    the variance of the direct effects set to zero \citep{kolesar12re}
    \begin{equation*}
      \h{\mathcal{H}}{ure}=
      \frac{1}{\hat{\Lambda}_{22}^{2}}\left(\hat{\Lambda}_{22}\hat{\Sigma}_{11}+
        \frac{(1-L/n )K/n}{(1-K/n-L/n)}
        (\hat{\Sigma}_{11}\hat{\Sigma}_{22}+\hat{\Sigma}_{12}^{2})    \right),
    \end{equation*}
where
\begin{align*}
  \hat{\Lambda}_{22}&= \begin{cases} S_{22}-\frac{K }{n}\h{\Omega}{22,ure} &
    \text{if $(\min\eig(S_{\perp}^{-1}S))\geq K/n$,} \\
    \frac{\h{\lambda}{re}}{\h{a}{re}'\h{\Omega}{re}^{-1}\h{a}{re}} &
    \text{otherwise.}
      \end{cases},\\
      \h{\Omega}{ure}&=\begin{cases}
        S_{\perp} & \text{if $(\min\eig(S_{\perp}^{-1}S))\geq K/n$,} \\
        \h{\Omega}{re} & \text{otherwise.}
      \end{cases},\\
      \h{\Sigma}{re} &= \h{\Gamma}{re}^{-1}\h{\Omega}{re}\h{\Gamma}{re}^{-1}, &
      \h{\Gamma}{re} &=\begin{pmatrix} 1& \hat{\beta}_{\mathtt{mbtsls}}\\
        0 &1
  \end{pmatrix}.
  % \h{\Sigma}{ure}&= {\h{\Gamma}{ure}}'\h{\Omega}{ure} \h{\Gamma}{ure}???&
  % \h{\Gamma}{ure}&=\begin{pmatrix} 1 & 0\\-\h{\beta}{ure}&1
  % \end{pmatrix}
\end{align*}

\todo[inline]{todo: compute UJIVE standard errors}

\paragraph{Many invalid instruments}
The estimator for the standard error of \texttt{mbtsls} is based on the Hessian
of the uncorrelated random effects likelihood,
    \begin{equation*}
      \h{\mathcal{H}}{ure}=
      \frac{1}{\hat{\Lambda}_{22}^{2}}\left(\hat{\Lambda}_{22}\hat{\Sigma}_{11}+
        \frac{(1-L/n )K/n}{(1-K/n-L/n)}
        (\hat{\Sigma}_{11}\hat{\Sigma}_{22}+\hat{\Sigma}_{12}^{2})
        +\hat{\Lambda}_{11}\hat{\Sigma}_{22}+
        \hat{\Lambda}_{11}\Lambda_{22}\cdot n/K
      \right),
    \end{equation*}
    where $\hat{\Lambda}_{22}$ and $\hat{\Sigma}_{22}$ are computed as before,
    and $\hat{\Lambda}_{11}=\max\left\{\hat{b}_{\mathtt{mbtsls}}'(S-\frac{K
        }{n}S_{\perp}) \hat{b}_{\mathtt{mbtsls}} ,0\right\},$


\todo[inline]{todo: compute UJIVE standard errors}


\section{Other outputs}
The command
\begin{lstlisting}
[betahat, se, stats] = ivreg(Y, T, Z, W);
\end{lstlisting}
returns a cell array \lstinline!stats! that contains additional statistics and
their names. In particular, it contains the first-stage $F$-statistic,
\begin{align*}
  \text{\lstinline!stats\{1,1\}!} &= \frac{\bt'\Hm{\bz_{\perp}}\bt}{K \cdot\bt'\Mm{\bz,\bw}\bt /(n-K-L)},&
  \text{\lstinline!stats\{1,2\}!} &= \text{\lstinline!'F'!},
\end{align*}
an estimator of the reduced-form covariance matrix
\begin{align*}
  \text{\lstinline!stats\{2,1\}!}&=
  \begin{pmatrix}
    \by&\bt
  \end{pmatrix}' \Mm{\bz,\bt} \begin{pmatrix}
    \by&\bt
  \end{pmatrix}/(n-K-L)&
  \text{\lstinline!stats\{2,2\}!} &= \text{\lstinline!'Omega'!},
\end{align*}
an estimator of the covariance matrix of the reduced form coefficients,
$\Xi=\Pi'\bz_{\perp}'\bz_{\perp}\Pi/n$ (see \citet{kolesar12late}), where $\Pi$
is the coefficient on $Z_{i}$ in the linear predictor
$\LP{(Y_{i},T_{i})}{Z_{i},W_{i}}$:
\begin{align*}
  \text{\lstinline!stats\{3,1\}!}&=
  \begin{pmatrix}
    \by&\bt
  \end{pmatrix}' \Hm{\bz_{\perp}} \begin{pmatrix}
    \by&\bt
  \end{pmatrix}/n-K/n\cdot     \text{\lstinline!stats\{2,1\}!}&
  \text{\lstinline!stats\{3,2\}!} &= \text{\lstinline!'Xi'!},
\end{align*}
The estimator is consistent under the many-instrument asymptotic sequence and
homoscedasticity.



\begin{appendices}
  The appendix derives the expressions for different estimators an gives
  informal proofs of consistency of estimators of the asymptotic variance.

  Let $\bx=(\bt,\bw)$ denote the full matrix of covariates in the structural
  equation.

\section{Estimators}
If an estimator of $(\beta,\psi')'$ can be written as
$(\hat{\bx}'\bx)^{-1}\hat{\bx}'\by$, where $\hat{\bx}=(\hat{\bt},\bw)$, we
obtain:
  \begin{align}\label{eq:betapsi}
    \begin{pmatrix}
      \hat{\beta}\\\hat{\psi}
    \end{pmatrix}=
    \begin{pmatrix}
      \hat{\bt}'\bt& \hat{\bt}'\bw\\
      \bw'\bt&\bw'\bw
    \end{pmatrix}^{-1}
    \begin{pmatrix}
      \hat{\bt}'\by\\
      \bw'\by
    \end{pmatrix}
    =
    \begin{pmatrix}
      (\hat{\bt}'\Mm{\bw}\bt)^{-1}\hat{\bt}'\Mm{\bw}\by\\
      (\bw'\bw)^{-1}\bw'(\by-\bt\hat{\beta}),
    \end{pmatrix}
  \end{align}
  so that letting $\hat{\bp}=\Mm{\bw}\hat{\bt}$, an estimator of $\beta$ is
  given by Equation \eqref{eq:hatbeta}. Now:
  \begin{align*}
    \hat{\bx}_{\mathtt{ols}}&=\bx,\\
    \hat{\bx}_{\mathtt{tsls}}&=\Hm{\bz,\bw}\bx=
    \begin{pmatrix}
      \Hm{\bz,\bw}\bt& \bw,
    \end{pmatrix}
  \end{align*}
  which, using the identity
  \begin{equation}\label{eq:Hzw}
    \Mm{\bw}\Hm{\bz,\bw}=\Mm{\bw}(\Hm{\bz_{\perp}}+\Hm{\bw})=\Hm{\bz_{\perp}},
  \end{equation}
  leads to the expressions for the \texttt{ols} and \texttt{tsls} estimators of
  $\beta$. The espressions for \texttt{liml}, \texttt{mbtsls}, and \texttt{jive}
  follow similarly. The expression for \texttt{ujive} and \texttt{rtsls} are
  derived in \citet{kolesar12late}.

\section{Standard Errors}

\paragraph{Homoscedasticity}
The homoscedastic case strengthens the moment condition \eqref{eq:id} to
\begin{align*}
  \CE{\epsilon_{i}}{Z_{i},W_{i}}&=0&
  \CE{\epsilon_{i}^{2}}{Z_{i},W_{i}}&=\E[\epsilon^{2}_{i}]=\sigma^{2}_{\epsilon}
\end{align*}
An estimator of variance of the structural error is then given by
$\hat{\sigma}^{2}_{\epsilon}=\hat{\epsilon}'\hat{\epsilon}/n$, where
$\hat{\epsilon}=\by-\bt\hat{\beta}-\bw'\hat{\psi}$. If an estimator admits the
representation \eqref{eq:betapsi}, then $\hat{\epsilon}$ can be written as
$\hat{\epsilon}=\Mm{\bw}(\by-\bt\hat{\beta})$, which leads to the expression for
$\hat{\sigma}^{2}_{\epsilon}$ given by Equation \eqref{eq:sigeps}.

The asymptotic variance of the \texttt{ols} estimator of the best linear
predictor, $(\E[X_{i}X_{i}']^{-1}\E[X_{i}'y_{i}])$, is given by
$\sigma_{\epsilon}^{2}\E[X_{i}X_{i}']^{-1}$. A the standard White estimator of
this variance is given by
\begin{equation*}
  \hat{\sigma}^{2}_{\epsilon}(\bx'\bx)^{-1}.
\end{equation*}
Since the (1,1) element of $(\bx'\bx)^{-1}$ is given by $(\bt\Mm{\bw}\bt)^{-1}$,
the expression for the \texttt{ols} standard error follows. The asymptotic
variance for \texttt{tsls}, \texttt{liml}, and \texttt{mbtsls} estimators of
$(\beta,\psi)$ is given by\todo{check the \citet{dm93} reference for asymptotic variances}\footnote{see \citet[Equation 5.24]{wooldridge02} and
\citet{dm93}.}
\begin{equation*}
  \sigma^{2}_{\epsilon}\left\{\E[X_{i}\tilde{Z}_{i}']
    \E[\tilde{Z}_{i}\tilde{Z}_{i}']^{-1}\E[\tilde{Z}_{i}X_{i}']\right\}^{-1},
\end{equation*}
where $\tilde{Z}_{i}=(Z_{i}',W_{i}')'$. This leads to the plug-in estimator
\begin{equation*}
  \hat{\sigma}^{2}_{\epsilon}(\hat{\beta})(\bx' \Hm{\bz,\bw}\bx)^{-1}.
\end{equation*}
Now, since
\begin{equation*}
  \begin{split}
    (\bx'\Hm{\bz,\bw}\bx)^{-1}&=
    \begin{pmatrix}
      \bt'\Hm{\bz,\bw}\bt & \bt'\bw\\
      \bw'\bt & \bw'\bw
    \end{pmatrix}^{-1}\\
    &=
    \begin{pmatrix}
      (\bt'\Hm{\bz_{\perp}}\bt)^{-1} & -  (\bt'\Hm{\bz_{\perp}}\bt)^{-1}\bt'\bw(\bw'\bw)^{-1}\\
      -(\bw'\bw)^{-1}\bw'\bt (\bt'\Hm{\bz_{\perp}}\bt)^{-1}& (\bw'\bw)^{-1}+
      (\bw'\bw)^{-1}\bw'\bt (\bt'\Hm{\bz_{\perp}}\bt)^{-1}\bt'\bw(\bw'\bw)^{-1}
    \end{pmatrix},
  \end{split}
\end{equation*}
the expression for the \texttt{tsls}, \texttt{liml}, and \texttt{mbtsls}
standard errors follows.

\lstinline!ivreg! uses the Stata estimator of the variance of \texttt{jive}
\citep{poi06}, given by
\begin{equation*}
    \hat{\sigma}^{2}_{\epsilon}(\hat{\beta})(\hat{\bx}'\bx)^{-1}\hat{\bx}'\hat{\bx}(\bx'\hat{\bx})^{-1},
\end{equation*}
where $\hat{\bx}=(\hat{\bt},\bw)$,
$\hat{\bt}=(\eye_{n}-\Dm{\bz,\bw})^{-1}(\Hm{\bz,\bw}-\Dm{\bz,\bw})\bt$.
Therefore,
\begin{equation*}
  (\hat{\bx}'\bx)^{-1}=
  % \begin{pmatrix}
  %   \hat{\bt}'\bt&\hat{\bt}'\bw\\
  %   \bw'\bt& \bw'\bw
  % \end{pmatrix}^{-1}=
  \begin{pmatrix}
    (\hat{\bp}'\bt)^{-1}&-
    (\hat{\bp}'\bt)^{-1}\hat{\bt}'\bw(\bw'\bw)^{-1}\\
    -(\bw'\bw)^{-1} \bw'\bt(\hat{\bp}'\bt)^{-1}& (\bw'\bw)^{-1}+(\bw'\bw)^{-1}
    \bw'\bt (\hat{\bp}'\bt)^{-1} \hat{\bt}'\bw(\bw'\bw)^{-1}
  \end{pmatrix},
\end{equation*}
where $\hat{\bp}=\Mm{\bw}\hat{\bt}=\hat{\bp}_{\mathtt{jive}}$ so that the (1,1)
element of the estimator of the variance evaluates as
\begin{equation*}
    \hat{\sigma}^{2}_{\epsilon}(\hat{\beta})(\hat{\bp}'\bt)^{-1}  (\hat{\bt}'\Mm{\bw}\hat{\bt})(\bt'\hat{\bp})^{-1}.
\end{equation*}
The expression for the \texttt{jive} standard error estimator follows.

\todo[inline]{justify UJIVE}

\paragraph{Heteroscedasticity}
Without the additional homoscedasticity restrictions on the structural error,
the asymptotic variance of the \texttt{ols} estimator is given by
$\E[X_{i}X_{i}']^{-1}\E[\epsilon_{i}^{2}X_{i}X_{i}']\E[X_{i}X_{i}']^{-1}$, and
the standard White estimator of the \texttt{ols} variance is given by
\begin{equation*}
  (\bx'\bx)^{-1}\sum_{i=1}^{n}\hat{\epsilon}_{i}^{2}X_{i}X_{i}'   (\bx'\bx)^{-1}
\end{equation*}
A robust estimator of the variance of
\texttt{tsls}, \texttt{liml}, and \texttt{mbtsls} is given by \citep[Equation
5.34]{wooldridge02}
\begin{equation*}
  (\hat{\bx}'\hat{\bx}')^{-1}
  \sum_{i=1}^{n}\hat{\epsilon}_{i}^{2}(\hat{\bx})_{i}(\hat{\bx})_{i}'
  (\hat{\bx}'\hat{\bx}')^{-1},
\end{equation*}
where $\hat{\bx}=(\Hm{\bz,\bw}\bt,\bw)$. For \texttt{jive}, I use the Stata
implementation \citep{poi06}
\begin{equation*}
  (\hat{\bx}'\bx)^{-1}\sum_{i}\hat{\epsilon}_{i}^{2}\hat{\bx}_{i}\hat{\bx}_{i}'
  (\bx'\hat{\bx})^{-1},
\end{equation*}
wherewhere $\hat{\bx}=(\hat{\bt},\bw)$,
$\hat{\bt}=(\eye_{n}-\Dm{\bz,\bw})^{-1}(\Hm{\bz,\bw}-\Dm{\bz,\bw})\bt$.All of these estimators have the form
\begin{equation}\label{eq:2}
  (\hat{\bx}'\bx)^{-1}\sum_{i}\hat{\epsilon}_{i}^{2}\hat{\bx}_{i}\hat{\bx}_{i}'
  (\bx'\hat{\bx})^{-1},
\end{equation}
where $\hat{\bx}=(\hat{\bt},\bw)$ for some $\hat{\bt}$. The (1,1) element of
\eqref{eq:2} can be written as
\begin{multline*}
  (\bt'\Mm{\bw}\hat{\bt})^{-1}
  \left[\sum_{i=1}^{n}\hat{\epsilon}_{i}^{2}\hat{\bt}_{i}^{2} -
    \sum_{i=1}^{n}\hat{\epsilon}_{i}^{2}\hat{\bt}_{i}W_{i}'\hat{\phi}
    -\hat{\phi}' \sum_{i=1}^{n}\hat{\epsilon}_{i}^{2}W_{i}\hat{\bt}_{i}
    +\hat{\phi}' \sum_{i=1}^{n}\hat{\epsilon}_{i}^{2}W_{i}W_{i}'
    \hat{\phi}\right]  (\hat{\bt}'\Mm{\bw}\bt)^{-1}  \\
  =(\bt_{\perp}'\bt_{\perp})^{-1} \sum_{i=1}^{n}\hat{\epsilon}_{i}^{2}
  \left(\hat{\bt}_{i}-W_{i}'\hat{\phi}\right)^{2}
  (\bt_{\perp}'\bt_{\perp})^{-1},
\end{multline*}
where $\hat{\phi}=(\bw'\bw)^{-1}\bw'\hat{\bt}$. Since
$(\Mm{\hat{\bt}})_{i}=\hat{\bt}_{i}-W_{i}'\hat{\phi}$, the expressions for the
standard errors given in the test follow. \todo[inline]{justify ujive}
\end{appendices}

\dotfill





\begin{align*}
  \begin{split}
    (\bx'\Hm{\bz,\bw}\bx)^{-1}\bx'\Hm{\bz,\bw}\by &=
    \begin{pmatrix}
      (\bt'\Hm{\bz_{\perp}}\bt)^{-1} \bt'\Hm{\bz_{\perp}}\by\\
      (\bw'\bw)^{-1} \bw'(\by-\bt (\bt'\Hm{\bz_{\perp}}\bt)^{-1}
      \bt'\Hm{\bz_{\perp}}\by)
    \end{pmatrix}
  \end{split}
\end{align*}

\SmallBib
\bibliography{/home/kolesarm/texmf/db/wi,/home/kolesarm/texmf/db/metrics,/home/kolesarm/texmf/db/books}



\end{document}
%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
